### 2025-11-05 sections/3-datasets.tex
down to 7 pages

\subsection{ANN-Benchmark Datasets}
\label{sec:datasets-and-benchmarks:ann-benchmark-datasets}

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
Table~\ref{tab:datasets:summary} summarizes the cardinality and dimensionality of each these datasets.
All benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.164-1-MANJARO.
The Rust compiler was Rust 1.83.0, and the Python interpreter version was 3.9.18.

@@

\subsection{Random Datasets and Synthetic Augmentation}
\label{sec:datasets-and-benchmarks:random-datasets}

In addition to the ANN-Benchmarks datasets, we evaluate synthetic augmentations generated as in Sec.~\ref{sec:methods:synthetic-data}, using $\epsilon=0.01$, and study scaling behavior as the multiplier (“Mult.” in Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random}) increases.
We also benchmark a random dataset at various cardinalities;
we use a random dataset with base cardinality of $10^6$ and dimensionality $128$, and refer to this dataset as ``Random.''
This benchmark isolates the effect of manifold structure—expected to be absent in purely random data—on CAKES algorithms.

\subsection{SILVA 18S}
\label{sec:datasets-and-benchmarks:silva-18s}

To demonstrate CAKES with an exotic distance function, we use the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}, which contains rRNA sequences from 2,224,640 genomes;
the longest sequence being 3,712 letters.
Holding out 1,000 random sequences as queries, we build the tree and perform $k$-NN using Levenshtein distance~\cite{levenshtein1966binary} on \emph{unaligned} sequences.
Although this dataset is a multiple sequence alignment of width 50,000, we do not use Hamming distance on the aligned strings so as to demonstrate CAKES's flexibility.
Under Hamming, the embedding dimension would be 50,000;
under Levenshtein, the dimensionality is 3,712.

\subsection{Radio ML}
\label{sec:datasets-and-benchmarks:radio-ml}

We also use the Radio-ML dataset~\cite{oshea2018radioml} under Dynamic Time Warping~\cite{muller2007dynamic} as the metric.
This dataset contains 24 modulation modes spanning 26 SNR levels from $-20$dB to $30$dB, with 4,096 samples per combination, yielding 2,555,504 samples.
Each sample is a 1,024-dimensional complex-valued vector (a time-series).
For the benchmarks, we use the 10dB SNR subset with 97,304 samples and a 1,000-sample hold-out set of queries.

\subsection{Local Fractal Dimension}
\label{sec:dayasets:lfd-of-datasets}

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks (Figure~\ref{fig:results:lfd-plots}).
In this section, when we discuss LFD, unless otherwise noted, we are referring to the 95th percentile from Figure~\ref{fig:results:lfd-plots}.

For Fashion-MNIST, Figure~\ref{fig:results:fashion-mnist-lfd}, we observe that until depth 5, LFD is less than 4. 
It then starts increasing, reaching a peak of about 6 near depth 20.
In comparison, Glove-25 has even lower LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}. 
In particular, Glove-25's LFD is less than 3 for all depths.
Figure~\ref{fig:results:sift-lfd} shows the LFD for Sift, which is higher relative to both Fashion-MNIST and Glove-25, increasing sharply to a peak of 9 around depth 10.
For Random, which has the same cardinality and dimensionality as Sift, LFD starts at 20 the root cluster, and all percentile values decrease linearly with depth until reaching the leaves of the tree.
The spread in LFD starts narrow for the first few depths and widens as depth increases.
Silva, Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD;
less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.
For the Radio-ML dataset, Figure~\ref{fig:results:radioml-lfd} LFD values show three distinct peaks around an LFD of 12 near depths of 8, 25 and 50.
Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
Within each of the three spans of depths, the character of Radio-ML is similar to that of Random.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets. The ‘random’ dataset is generated as in Sec.~\ref{sec:datasets-and-benchmarks:random-datasets}; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, plus min/max. Each cluster is counted by its cardinality so the curves reflect dataset-level distributions.}
    \label{fig:results:lfd-plots}
  \end{figure}

% Since the CAKES algorithms were designed to scale in time with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
% Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and Radio-ML.
% In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.

% For the Fashion-MNIST dataset we observe in Figure~\ref{fig:results:fashion-mnist-lfd} that until approximately depth 5, Fashion-MNIST's LFD is low (i.e., less than 4).
% It then starts increasing, reaching a peak of about 6 near depth 20, before decreasing to 1 at the maximum depth.

% Relative to Fashion-MNIST, Glove-25 has low LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}.
% In particular, Glove-25's LFD is less than 3 for all depths.

% Figure~\ref{fig:results:sift-lfd} shows the LFD by depth for Sift, which has higher LFD relative to Fashion-MNIST and Glove-25.
% It increases sharply to a peak of 9 around a depth of 10, after which it decreases smoothly until reaching the deepest leaves in the tree.

% We contrast the low LFD of Sift with that of the Random dataset which has the same cardinality and dimension. As shown in Figure~\ref{fig:results:random-lfd}, the character of this dataset is very different from the others.
% The LFD starts at 20 at depth 0 and all percentile lines decrease linearly with depth until reaching the leaves of the tree.
% The spread in LFD starts very small for the first few clusters and increases as depth increases.

% The Silva dataset, shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
% In particular, LFD is less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

% For the Radio-ML dataset, the LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
% Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
% Within each of the tree portions, this dataset has a character similar to that of the Random dataset.
% This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth.
% This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.

@@

We benchmark the three CAKES algorithms against a na\"ive linear search implementation in Rust.
We also benchmark against three state-of-the-art similarity search algorithms: HNSW, ANNOY, and FAISS-IVF in Python.
We verified that our linear search produces the same results as in the ground truth of the ANN-Benchmarks suite for Fashion-MNIST, Glove-25 and Sift datasets.
We then used this implementation to find and store the ground-truth for the augmented versions of those datasets, and use it to calculate recall for all algorithms.
We plot the results of these benchmarks in Figure~\ref{fig:results:scaling-plots}.

>>>>

\subsection{Datasets}
\label{sec:datasets-and-benchmarks:ann-benchmark-datasets}

Table~\ref{tab:datasets:summary} summarizes the cardinality and dimensionality of each dataset we use.
These include the Fashion-MNIST, Glove-25, and Sift datasets from the ANN-Benchmarks suite~\cite{aumuller2020ann}, as well as a Random dataset we generated (using the same cardinality and dimensionality as Sift) as an anti-example to demonstrate the effect of manifold structure on CAKES performance.
Thus, Random shows that CAKES is designed to exploit manifold structure when it exists, and should not be used on datasets that do not obey the manifold hypothesis.
The Glove-25 dataset uses the Cosine distance function, while the other three datasets use the Euclidean distance metric.
All benchmarks using these datasets were conducted on an r6i.16xlarge instance on Amazon Web Services (AWS), with 64 vCPUs and 512 GiB of RAM.
The Rust compiler was Rust 1.83.0 and the Python interpreter version was 3.9.18.

We also include benchmarks on the SilVA-18S rRNA and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and use Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1-MANJARO with Rust compiler version 1.83.0.

@@

\subsection{Scaling Experiments}
\label{sec:datasets-and-benchmarks:scaling-experiments}

In order to study scaling behavior as cardinality increases, we synthetically augment the ANN-Benchmarks datasets by generating new points by applying small random perturbations to existing points in the original dataset.
Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random} show the results as the cardinality increases, with ``Mult.'' indicating the multiplier by which the cardinality was augmented.
Since the SilVA-18S and Radio-ML datasets are already large and are not vector datasets, we do not augment them;
instead, we subsample them to lower cardinalities for the scaling experiments.

\subsection{Local Fractal Dimension}
\label{sec:datasets:lfd-of-datasets}

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine, in Figure~\ref{fig:results:lfd-plots}, the LFD of each dataset we used for benchmarks.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD indicating the presence of manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process that imparts some manifold structure at three distinct radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, the throughput is averaged over all queries, and recall is computed using a na\"ive linear scan.

HNSW, ANNOY, and FAISS have various hyper-parameters that can be tuned to trade off throughput, recall, and indexing time.
We allow for a hyper-parameter search (in the ranges recommended by the documentation accompanying the respective libraries) to find the highest throughput configuration that achieves perfect recall (or as close to it as possible) for each dataset and cardinality.
To make the comparison fair, we also allow CAKES to select the optimal algorithm among the three CAKES algorithms for each dataset and cardinality by using a small number of the centers of clusters at low depth as representative queries.

### 2025-11-05 sections/4-results.tex
down to 7 pages

For every dataset and distance funciton listed in Table~\ref{tab:datasets:summary}, we benchmark $k$-NN search using $k=10$.
Smaller values of $k$, as described in Section~\ref{sec:introduction}, would be too sensitive to local perturbations, while larger values can capture too much of the global structure.

In Figure~\ref{fig:results:scaling-plots}, we show how the throughput of each algorithm scales with cardinality for each of the six datasets we examined.
To do this, we synthetically augmented the Fashion-MNIST, Glove-25, Sift, and Random datasets to create larger datasets using the process described in Section~\ref{sec:methods:synthetic-data}.
For Silva and RadioML, due to the massive sizes of these datasets and challenges in generating plausible augmentations, we took random sub-samples ranging up to the entirety of the dataset to examine how performance scales with cardinality.
These plots also illustrate how the CAKES algorithms compare to na\"{i}ve linear search.

Tables~\ref{tab:results:qps-and-recall-fmn},~\ref{tab:results:qps-and-recall-glove},~\ref{tab:results:qps-and-recall-sift} and~\ref{tab:results:qps-and-recall-random} compare the performance (throughput and recall) of the CAKES algorithms against state-of-the-art algorithms on the Fashion-MNIST, Glove-25, Sift and Random datasets.
In particular, we examine performance of HNSW, ANNOY and FAISS-IVF on each of those datasets as well as synthetically augmented versions of the datasets to isolate the effect of dataset size on performance.
We did not perform similar benchmarks on the Silva and RadioML datasets because HNSW, ANNOY and FAISS support neither the required distance functions nor, in the case of RadioML, complex-valued data.
For HNSW, ANNOY, and FAISS-IVF, we allow for a hyper-parameter search to tune their index for maximum recall.
For CAKES, we build the tree and use our auto-tuning approach (see Section~\ref{sec:methods:auto-tuning}) to select the fastest algorithm for each dataset and cardinality.

Though the plots in Figure~\ref{fig:results:scaling-plots} present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the fastest CAKES algorithm at that dataset and cardinality only.

Throughput is measured in queries per second.
A recall value of $1.000*$ denotes imperfect recall that rounds to $1.000$.
In each of Tables \ref{tab:results:qps-and-recall-fmn},\ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, we observe that while recall for CAKES does \emph{not} degrade with cardinality, recall for
HNSW and ANNOY does degrade with cardinality. CAKES exhibits perfect recall on the Fashion-MNIST and Sift datasets, and near-perfect recall on the Glove-25 dataset (which uses cosine distance).

% Finally, we test our intuitions about unbalanced clustering being better for search
% than balanced clustering (see Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations} and Figure~\ref{fig:results:distance-counts}).

% \subsection{Local Fractal Dimension of Datasets}
% \label{sec:results:lfd-of-datasets}

% Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
% Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and Radio-ML.
% In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.
% This is because our algorithms scale exponentially in the LFD, a

% The Fashion-MNIST dataset has an embedding dimension of 784 and uses the Euclidean distance metric.
% In Figure~\ref{fig:results:fashion-mnist-lfd} we observe that until approximately depth 5, Fashion-MNIST's LFD is low (i.e., less than 4).
% It then starts increasing, reaching a peak of about 6 near depth 20, before decreasing to 1 at the maximum depth.

% The Glove-25 dataset has an embedding dimension of 25 and uses the cosine distance function which, notably, is not a metric.
% Relative to Fashion-MNIST, Glove-25 has low LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}.
% All percentile lines for Glove-25 are flatter and lower, indicating that the LFD is lower across the entire dataset, and that the LFD does not vary as much by depth.
% In particular, Glove-25's LFD is less than 3 for all depths.

% The Sift dataset has an embedding dimension of 128 and uses the Euclidean distance metric.
% Figure~\ref{fig:results:sift-lfd} shows the LFD by depth for Sift, which has higher LFD relative to Fashion-MNIST and Glove-25.
% It increases sharply to a peak of 9 around a depth of 10.
% It then decreases smoothly until reaching the deepest leaves in the tree.

% We generated the Random dataset to have the same cardinality and dimensionality as Sift.
% We used a uniform distribution in a 128-dimensional unit-hypercube to generate the points and the Euclidean metric to measure distances among them.
% Figure~\ref{fig:results:random-lfd} shows that the character of this dataset is significantly different from the others.
% The LFD starts at 20 at depth 0 and all percentile lines decrease linearly with depth until reaching the leaves of the tree.
% The spread in LFD starts very small for the first few clusters and increases as depth increases.
% The LFD of approximately 20 for the root cluster $\mathcal{R}$ is what we expect for this random dataset.
% To elaborate, the distribution of points in such a dataset should reflect the curse of dimensionality, i.e.,\,the fact that in high dimensional spaces, the minimum and maximum pairwise distances between any two points are approximately equal.
% As a result, $\mathcal{R}$'s radius $r$, which reflects the maximum distance between the center $c$ and any other point, should not differ significantly from the distance between the center and its closest point.
% A consequence of this is that, with high probability, for every point in $\mathcal{R}$, its distance from $c$ is greater than $\tfrac{r}{2}$;
% in other words, $B(c, \tfrac{r}{2})$ contains only $c$ while $B(c, r)$ contains the entire dataset.
% Given our definition of LFD in Equation~\ref{eq:methods:lfd-half}, this means that the LFD of $\mathcal{R}$ is approximately $\log_2(\frac{|X|}{1}) = \log_2(1,000,000) \approx 20$, which is what we observe in Figure~\ref{fig:results:random-lfd}.
% Theoretically, the LFD of this dataset should be 128, i.e.\, it should be the same as the embedding dimension.
% This reflects the difference between how we empirically measure the LFD and the value we would expect.
% With sample sizes larger than 1,000,000, we would expect the LFD to approach 128 until we have sampled $2^{128}$ points, at which point the LFD would be 128 and would stay at 128 for even larger sample sizes.
% Unfortunately, such a large sample is practically impossible to generate.

% The Silva-18S dataset consists of genomic sequences whose unaligned lengths are at-most 3,712.
% As such the embedding dimension is 3,712, though as discussed previously, the embedding dimension would be 50,000 in a multiple sequence alignment.
% We use the Levenshtein edit distance (a metric) to measure distances between sequences.
% This dataset, as shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
% In particular, LFD is less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

% The Radio-ML dataset consists of measurements of radio-frequency signals using 1,024 dimensional complex-valued vectors.
% We use the Dynamic Time Warping distance metric on this dataset.
% This dataset is synthetic~\cite{oshea2018radioml} but uses a far more elaborate generation process than our Random dataset.
% The LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
% Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
% Within each of the tree portions, this dataset has a character very similar to that of the Random dataset.
% This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth.
% This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.

@@

\begin{figure}[h]
  \vskip -0.2in
  \centering
  \subfloat[Fashion-MNIST]{
    \includegraphics[width=0.48\columnwidth]{plots/fashion-mnist_PermutedBall_10_throughput.png}
    \label{fig:results:fashion-mnist-scaling}
  }
  \subfloat[Glove-25]{
    \includegraphics[width=0.48\columnwidth]{plots/glove-25_PermutedBall_10_throughput.png}
    \label{fig:results:glove-25-scaling}
  }

  \vskip -0.1in
  \subfloat[Sift]{
    \includegraphics[width=0.48\columnwidth]{plots/sift_PermutedBall_10_throughput.png}
    \label{fig:results:sift-scaling}
  }
  \subfloat[Random dataset]{
    \includegraphics[width=0.48\columnwidth]{plots/random_PermutedBall_10_throughput.png}
    \label{fig:results:random-scaling}
  }

  \vskip -0.1in
  \subfloat[Silva]{
    \includegraphics[width=0.48\columnwidth]{plots/silva-SSU-Ref_PermutedBall_10_throughput.png}
    \label{fig:results:silva-scaling}
  }
  \subfloat[RadioML at SnR = 10dB.]{
    \includegraphics[width=0.48\columnwidth]{plots/radio-ml_Ball_10_throughput.png}
    \label{fig:results:radioml-scaling}
  }

@@

  \caption{Throughput across six datasets, including a randomly-generated dataset.
  Each plot shows throughput (queries per second; higher is better) versus dataset cardinality. For Fashion-MNIST, Glove-25, and Sift, CAKES becomes faster than linear search as cardinality grows, with the crossover point varying by dataset. For Fashion-MNIST and Glove-25, Depth-First Sieve is consistently fastest. For Sift, Repeated $\rho$-NN is fastest at small cardinalities, while Depth-First Sieve is fastest at large cardinalities. For Silva, throughput for all algorithms initially decreases approximately linearly with cardinality and then levels off at higher cardinalities; Depth-First Sieve is consistently fastest. For Radio-ML and Random, all CAKES variants are slower than na\"{i}ve linear search, and their throughput decreases linearly with cardinality. HNSW and ANNOY are the fastest algorithms on all four datasets we benchmarked them on, but their recall degrades quickly as cardinality increases on all datasets; on Random, their recall is near zero.}
  \label{fig:results:scaling-plots}
\end{figure}

@@

% \begin{figure}
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnRepeatedRnn_10_throughput.png}
%         \subcaption{Repeated $\rho$-NN}
%         \label{fig:results:fashion-mnist-counts-throughput}
%     \end{subfigure}%
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnRepeatedRnn_10_counts.png}
%         \subcaption{Repeated $\rho$-NN}
%         \label{fig:results:glove-25-counts-counts}
%     \end{subfigure}%
%     \\
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnBreadthFirst_10_throughput.png}
%         \subcaption{Breadth First Sieve}
%         \label{fig:results:sift-counts-throughput}
%     \end{subfigure}%
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnBreadthFirst_10_counts.png}
%         \subcaption{Breadth First Sieve}
%         \label{fig:results:random-counts-counts}
%     \end{subfigure}%
%     \\
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnDepthFirst_10_throughput.png}
%         \subcaption{Depth First Sieve}
%         \label{fig:results:silva-counts-throughput}
%     \end{subfigure}%
%     \begin{subfigure}[b]{0.5\textwidth}
%         \includegraphics[width=0.99\textwidth]{images/distance_counts/fashion-mnist_KnnDepthFirst_10_counts.png}
%         \subcaption{Depth First Sieve}
%         \label{fig:results:radioml-counts-counts}
%     \end{subfigure}%
%     \\
%     \begin{subfigure}[b]{0.94\textwidth}
%         \centering
%         \includegraphics[width=0.6\textwidth]{images/distance_counts/legend.png}
%         \label{fig:results:counts-legend}
%     \end{subfigure}%
%     \caption{Number of distance computations across four clustering strategies and three search algorithms on the Fashion-MNIST dataset.
%     Adding the instrumentation to count the number of distance computations had the side-effect of significantly slowing down the search algorithms compared to those reported in Figure~\ref{fig:results:scaling-plots}.
%     The left column shows the throughput in queries per second, while the right column shows the mean number of distance computations per query.
%     The x-axis represents increasing cardinality of the dataset.}
%     \label{fig:results:distance-counts}
% \end{figure}

\subsection{Clustering Strategies}
\label{sec:results:clustering-strategies-and-number-of-distance-computations}

Beyond scaling behavior, we examine how clustering strategy affects search performance.
We consider four strategies—the Cartesian product of balanced vs.\ unbalanced clustering and the presence vs.\ absence of depth-first reordering (Sec.~\ref{sec:methods:clustering:depth-first-reordering}).
We instrument CAKES to count distance computations during search;
this adds substantial overhead and would not be used in real-world applications.

Figure~\ref{fig:results:distance-counts} compares \textit{Ball}, \textit{BalancedBall}, \textit{PermutedBall}, and \textit{PermutedBalancedBall} (``Permuted'' denotes depth-first reordering) for all three algorithms on Fashion-MNIST.
Balanced clustering is strictly worse in throughput and has poorer asymptotics: algorithms such as Depth-First Sieve that appear nearly constant-time in Fig.~\ref{fig:results:scaling-plots} (with \textit{PermutedBall}) no longer do so under balanced clustering.
On this dataset, depth-first reordering has little effect on throughput; it is primarily intended to improve memory consumption.

\begin{figure}[h]
  \centering
  \subfloat[Repeated $\rho$-NN]{
  \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnRepeatedRnn_10_throughput.png}
  \label{fig:results:fashion-mnist-rnn-throughput}
  }
  \subfloat[Repeated $\rho$-NN]{
    \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnRepeatedRnn_10_counts.png}    
    \label{fig:results:fashion-mnist-rnn-counts}
  }

  \subfloat[Breadth First Sieve]{
    \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnBreadthFirst_10_throughput.png}
    \label{fig:results:fashionmnist-bfs-throughput}
    }
  \subfloat[Breadth First Sieve]{
    \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnBreadthFirst_10_counts.png}
    \label{fig:results:fashionmnist-bfs-counts}
    }

  \subfloat[Depth First Sieve]{
    \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnDepthFirst_10_throughput.png}
    \label{fig:results:fashionmnist-dfs-throughput}
    }
  \subfloat[Depth First Sieve]{
    \includegraphics[width=0.48\columnwidth]{images/distance_counts/fashion-mnist_KnnDepthFirst_10_counts.png}
    \label{fig:results:fashionmnist-dfs-counts}
  }

  \vspace{2pt}
  \includegraphics[width=0.75\columnwidth]{images/distance_counts/legend.png}

  \caption{Number of distance computations across four clustering strategies and three search algorithms on the Fashion-MNIST dataset.
  Adding the instrumentation to count the number of distance computations had the side-effect of significantly slowing down the search algorithms compared to those reported in Figure~\ref{fig:results:scaling-plots}.
  The left column shows the throughput in queries per second, while the right column shows the mean number of distance computations per query.
  The x-axis represents increasing cardinality of the dataset.}
  \label{fig:results:distance-counts}
\end{figure}

>>>>

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} summarize the throughput and recall of CAKES against state-of-the-art $k$-NN search algorithms on the Fashion-MNIST, Glove-25, Sift, and Random datasets respectively.
Throughput is measured in queries per second (QPS), and a recall of $1.000*$ indicates imperfect recall that rounds to $1.000$.

We see that CAKES has perfect recall on all datasets using the Euclidean distance metric, and near-perfect recall on the Glove-25 dataset using Cosine distance.
This is because Cosine distance violates the triangle inequality and is not a metric.
The throughput of CAKES drops slightly for the first two multipliers, but then remains relatively constant as cardinality increases, demonstrating the scalability of CAKES on manifold-structured data.
In contrast, the recall of HNSW and ANNOY degrades significantly as cardinality increases, even though their throughput remains high.
FAISS-IVF keeps high recall but its throughput degrades linearly with cardinality.
This indicates that these approximate algorithms are unable to maintain both high recall and high throughput as dataset sizes increase, even on manifold-structured data.

For the Random dataset, CAKES again attains perfect recall at all cardinalities, while HNSW and ANNOY exhibit near-zero recall, and FAISS-IVF attains perfect recall but with very low throughput.

On the Silva dataset (Fig.~\ref{fig:results:silva-scaling}), all algorithms initially exhibit throughput that decreases approximately linearly with cardinality, but then level off at higher cardinalities.
CAKES's Depth-First Sieve is consistently the fastest algorithm on this dataset.

On the Radio-ML dataset (Fig.~\ref{fig:results:radioml-scaling}), all CAKES variants are slower than na\"{i}ve linear search, and their throughput decreases linearly with cardinality.
This is expected, as the Radio-ML dataset is a randomly generated dataset and exhibits little manifold structure under the DTW distance metric.

@@

\begin{figure}[h]
  \vskip -0.2in
  \centering
  \subfloat[Silva]{
    \includegraphics[width=0.48\columnwidth]{plots/silva-SSU-Ref_PermutedBall_10_throughput.png}
    \label{fig:results:silva-scaling}
  }
  \subfloat[RadioML at SnR = 10dB.]{
    \includegraphics[width=0.48\columnwidth]{plots/radio-ml_Ball_10_throughput.png}
    \label{fig:results:radioml-scaling}
  }

@@

  \caption{Throughput vs. dataset cardinality on Silva-18S and Radio-ML.}
  \label{fig:results:scaling-plots}
\end{figure}

@@
